Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using Multiple Automatically Generated Templates

### 2013-09-01 paper.Rnw
Some of Julie's edits

One methodology that can be used to mitigate these sources of errors involves
the use of multiple manually segmented atlases and probabilistic segmentation
techniques, such as those found in the FreeSurfer package \citep{Fischl2002}.
FreeSurfer uses a probabilistic atlas of anatomical and tissue classes along
with spatial constraints for class labels encoded using a Markov random field
model to segment the entire brain.

@@

To the best of our knowledge there are only two automated segmentation methods
that attempt to define the hippocampus and its subfields using standard
T1-weighted data. The first is included with the FreeSurfer package
\citep{VanLeemput2009}.  This work is limited as it omits the tail of the
hippocampus and the segmentation protocol has yet to be fully validated.
Nonetheless, it demonstrates that the applicability of hippocampal subfield
segmentation using data from 10 subjects. In the second method,
\citet{Yushkevich2009} hippocampal subfields were acquired and labelled on
high-resolution MRI data from post-mortem medial temporal lobe samples. Using
nonlinear registration guided by manually derived hippocampus masks and specific
landmarks, the authors demonstrated accurate parcellation of hippocampal
subfields in unlabelled MRI volumes.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and SNT \gs segmentations from the ADNI1 dataset. This form of
cross-validation allows us to rigorously validate a large number of parameter
settings of \mb (atlas and template library sizes, registration algorithm, and
label fusion method) and select the best parameters to use in subsequent
experiments.

\paragraph{\adnidataset dataset} 
Data used in the preparation of this article were obtained from the Alzheimer’s
Disease Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). The ADNI
was launched in 2003 by the National Institute on Aging (NIA), the National
Institute of Biomedical Imaging and Bioengineering (NIBIB), the Food and Drug
Administration (FDA), private pharmaceutical companies and non-profit
organizations, as a $60 million, 5-year public- private partnership. The primary
goal of ADNI has been to test whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological markers, and clinical and
neuropsychological assessment can be combined to measure the progression of mild
cognitive impairment (MCI) and early Alzheimer’s disease (AD).  Determination of
sensitive and specific markers of very early AD progression is intended to aid
researchers and clinicians to develop new treatments and monitor their
effectiveness, as well as lessen the time and cost of clinical trials.

The Principal Investigator of this initiative is Michael W. Weiner, MD, VA
Medical Center and University of California – San Francisco. ADNI is the result
of efforts of many co- investigators from a broad range of academic institutions
and private corporations, and subjects have been recruited from over 50 sites
across the U.S. and Canada. The initial goal of ADNI was to recruit 800 subjects
but ADNI has been followed by ADNI-GO and ADNI-2.  To date these three protocols
have recruited over 1500 adults, ages 55 to 90, to participate in the research,
consisting of cognitively normal older individuals, people with early or late
MCI, and people with early AD. The follow up duration of each group is specified
in the protocols for ADNI-1, ADNI-2 and ADNI-GO. Subjects originally recruited
for ADNI-1 and ADNI-GO had the option to be followed in ADNI-2. For up-to-date
information, see www.adni-info.org.

@@

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.

@@

In this experiment, our dataset consists of 69 1.5T images and corresponding \gs
segmentations. In each validation round, the dataset is partitioned into a
training set consisting of images and their \gs segmentations to be used as an
atlas library, and a validation set consisting of the remaining images segmented
by both \mb and multi-atlas. The computed segmentations were compared to the
\gs segmentations (see below).

A total of ten validation rounds were performed on each subject in the dataset,
over each combination of parameter settings. The parameter settings
explored are: atlas library size (1-9), template library size (1-20),
registration method (\ants or \animal), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote).  A total of $10 \times 69 \times 9 \times 20 \times 2
\times 3 = \num{7452000}$ validation rounds were conducted, resulting in a total
of $\num{1490400}$ segmentations analysed.

@@

The \animal algorithm carries out image registration in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller full width at half maximum (FWHM). The final transformation is a set of
local translations defined on a bed of equally spaced nodes that were estimated
through the optimization of the correlation coefficient.  For the purposes of
this work we used the regularization parameters optimized in
\citet{Robbins2004}, displayed in Table \ref{tab:ANIMAL-params}.

@@

\paragraph{Label fusion methods}
Label fusion is a term given to the process of combining the information from
several candidate labels for an intensity image into a single labelling.  In
this experiment we explore three fusion methods. 
\begin{description}
  \item[Voxel-wise Majority Vote]
  Labels are propagated from all template library images to a target.  Each
  output voxel is given the most frequent label at that voxel location amongst
  all candidate labels.

@@

  \item[Normalised Mutual Information Weighted Majority Vote]
  This method is similar to cross-correlation weighted voting except that image
  similarity is calculated by the normalised mutual information score over the
  region of interest\citep{Studholme2001}.  The {\tt itk\_similarity} utility
  from the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is
  used to calculate the normalised mutual information measure between to images.
\end{description}

@@

\paragraph{Winterburn Atlases dataset} 
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
segmentations of five in-vivo $300\mu$ isotropic T1-weighted MR images. The
segmentations include subfield segmentations for the cornu ammonis (CA) 1, CA4,
dentate gyrus, subiculum, and CA 2 and 3 combined. Subjects in the Winterburn
atlases range in age from 29-57 years (mean age of 37), and include two males
and three females.

@@

In part 2a of this experiment, we have modified the LOOCV experimental
design in order to compare \mb segmentation accuracy with the accuracy of image
resampling error. The five $0.3mm$-isotropic voxel Winterburn atlas images and
labels are used as the atlas library, and LOOCV is conducted using the five Winterburn
atlas images subsampled (using trilinear interpolation) to
0.9mm-isotropic voxel resolution (referred to as the {\em Subsampled} dataset)
as input subjects. Subsampling the target images allows us to assess \mb in a
typical segmentation scenario (high-resolution atlases and lower-resolution
subjects).  The template library is composed of all five subject images, plus an
additional set of 3T T1 images (0.9mm-isotropic voxels) of healthy subjects
acquired separately (Table \ref{tab:WAval-healthy-demographics}). The optimal
size of template library, registration method, and label fusion method found in
Experiment 1 are used.  As with traditional LOOCV, each subject image is
segmented by \mb with that subject's image excluded from the atlas library.  In
part 2b of this experiment, we carry out a similar validation in which the
subject images are separately acquired T1 BRAVO images of four of the five
Winterburn atlas subjects (referred to as the {\em BRAVO} dataset).  Figure
\ref{fig:LOOCV-schematic} illustrates the experimental set up of both parts of
this experiment.

@@

\paragraph{Evaluation method}  
To assess the \mb LOOCV segmentations we compute the relative percent error in 
hippocampal volume with the full resolution Winterburn atlas 
segmentations. In addition, by computing the relative error in volume of the
Winterburn atlas labels resampled (with nearest-neighbour interpolation) to
0.9mm-isotropic voxels, we obtain a baseline error to assess against.

@@

To validate that \mb works effectively in the context of other neurological
disorders, in this experiment we use the Winterburn atlases to predict the
hippocampal segmentations of patients with Schizophrenia. The resulting
segmentations are assessed for quality by comparison with expert manual
segmentations.

@@

Manual segmentation of each subject hippocampus is produced following a
validated segmentation protocol \citep{Pruessner2000}.

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2).  All images from the \FEPdataset dataset are segmented by \mb.  The 
optimal size of template library, registration method, and label fusion method 
found in Experiment 1 are used.

\paragraph{Evaluation method}
The Pruessner and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated \citep{Winterburn2013}. This
difference poses a problem for evaluation by measuring overlap.  That is, since
different protocols will necessarily produce segmentations that do not perfectly
overlap, the degree of overlap cannot be solely used to compare segmentation
methods using different protocols. In place of an overlap metric, we can assess
the degree of correlation in average hippocampal volume of the subjects produced
by each method. Specifically, a Pearson correlation is used.

@@

To validate \mb segmentation quality with respect to other established
automated methods, \mb was applied to large dataset from the ADNI project and
the resulting segmentations were compared to those produced by FreeSurfer, FSL,
MAPER and the SNT \gs segmentations.

\paragraph{ADNI1 dataset revisited} 
The {\em \adnidataset} standardized dataset contains $1919$ images in total.
SNT and FreeSurfer hippocampal volumes for a subset of images were provided by
ADNI, along with quality control data for each FreeSurfer segmentation
(guidelines described in \citep{Hartig2010}).  Clinical and demographic data for
the entire \adnidataset dataset are shown in Table \ref{tab:ADNI1-Dataset-Demographics}.

@@

One author (MP) performed visual quality inspection for MAGeT and FSL
segmentations using similar quality control guidelines (if either hippocampus
was under or over segmented by greater than 10mm in three or more slices then
the segmentation did not pass).  Only images meeting the conditions of having
segmentations from all methods (SNT, MAPER, FreeSurfer, FSL, and \mb) and also
passing quality control inspection were included in the analysis.

\paragraph{Evaluation method}
As in Experiment 3, the SNT \gs and Winterburn hippocampal segmentation
protocols differ in the neuroanatomical features delineated, and so we 
assessed \mb by the degree of (Pearson) correlation of average hippocampal volume
across subjects. We also computed the correlation in hippocampal volume between
the existing, established automated segmentation methods: FSL, FreeSurfer, and
MAPER. Additionally, we evaluate the volume-related fixed and proportional
biases in all segmentation methods using Bland-Altman plots \citep{Bland1986}.

@@

In this experiment we conducted 10 rounds of \mb and multi-atlas segmentation of
of 69 subjects using atlas and template library sizes varying from 1-9 and 1-20
images respectively, two registration algorithms (\ants or \animal), and three
label fusion techniques (unweighted, cross-correlation, and normalised mutual
information weighted majority vote).  Hippocampal \mb-based segmentations using
both \animal and \ants registration algorithm demonstrate good overlap with SNT
\gs segmentations (maximum mean DSC of \Sexpr{round(mb.best.kappa, digits=3)}
when using 9 atlases, 19 templates, ANTS registration, and majority vote label
fusion); Figure \ref{fig:ADNI1-xval-k-mean}).  Qualitatively, both \animal and
\ants-based segmentations demonstrate trend overlap accuracy that increases with
the size of atlas library and template library. Improvement in accuracy plateaus
with template libraries larger than ten images.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<fig:ADNI1-xval-k-mean, cache=TRUE, dependson='ADNI1-xval-prep', fig.width=7, fig.height=7>>=
stderr <- function(x) sqrt(var(x)/length(x))
ggplot(subset(all_data_mean, (templates.mb*atlases) %% 2 == 1),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,
               fun.ymin=function (x) mean(x) - stderr(x),
               fun.ymax=function (x) mean(x) + stderr(x), 
               geom='errorbar',
               aes(y=k.mb, weight=1, group=as.factor(atlases)), width=0.5) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Number of Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  opts(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text = element_text(size = 8))
@
  \caption{Mean Dice's Similarity Coefficient of \mb segmentations with 
  SNT \gs segmentations for 69 ADNI1 subjects vs. atlas and template library
  size, registration algorithm, and label fusion method. Points above zero
  indicate \mb parameter settings yielding a higher mean DSC score than
  multi-atlas segmentation using the same number of atlases. Error bars
  indictate standard error.
  \label{fig:ADNI1-xval-k-mean}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: "increase" in mean kappa over multi-atlas
\begin{figure}
<<fig:ADN11-xval-k-diff,cache=TRUE,dependson='ADNI1-xval-prep'>>=
ggplot(subset(all_data_mean, reg_method=="ANTS" & 
                method.mb=="Majority Vote" & (templates.mb*atlases) %% 2 == 1), 
       aes(x=templates.mb, y=k_diff, colour=as.factor(atlases))) + 
  #facet_grid(reg_method~method.y) + 
  stat_summary(fun.data=mean_cl_boot,geom='errorbar',
               aes(y=k_diff,colour=as.factor(atlases), width=0.2)) +
  stat_summary(fun.y=mean, geom="point", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  stat_summary(fun.y=mean, geom="line", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  #stat_smooth(method=lm, formula=y~log(x)) + 
  geom_hline(aes(alpha=0.5, yintercept=0), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,2)) + scale_y_continuous(breaks=seq(-1,1,by=0.01)) + 
  scale_colour_hue(name="Number of Atlases") +
  scale_alpha_continuous(guide = "none") + 
  scale_linetype_discrete(guide = "none") + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference in mean DSC" ) +
  opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Difference in mean Dice's Similarity Coefficient of \mb and
  multi-atlas segmentations with \gs (SNT) segmentations for a range of atlas
  and template library sizes. Error bars indicate standard error.
  \label{fig:ADNI1-xval-k-diff}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

ggplot(subset(stats.tests, direction > 0 & stat == "Variance"), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  #facet_grid( . ~ stat) + 
  geom_line(size=1) +
  geom_point(size=3) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference of variance during validation (p-value)") +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  Variance of segmentation accuracy between \mb and multi-atlas segmentation 
  is computed for each subject across all ten rounds of validation. Shown on the 
  y-axis (scaled logarithmically) is the  p-value resulting from a t-test 
  comparing the distribution of variances at each parameter setting 
  (atlas/template library size). Only points where \mb mean variability is lower 
  than multi-atlas are shown.}
  \label{ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

One methodology that can be used to mitigate these sources of errors involves
the use of multiple manually segmented atlases and probabilistic segmentation
techniques, such as those found in the \freesurfer package \citep{Fischl2002}.
\freesurfer uses a probabilistic atlas of anatomical and tissue classes along
with spatial constraints for class labels encoded using a Markov random field
model to segment the entire brain.

@@

To the best of our knowledge there are only two automated segmentation methods
that attempt to define the hippocampus and its subfields using standard
T1-weighted data. The first is included with the \freesurfer package
\citep{VanLeemput2009}.  This work is limited as it omits the tail of the
hippocampus and the segmentation protocol has yet to be fully validated.
Nonetheless, it demonstrates that the applicability of hippocampal subfield
segmentation using data from 10 subjects. In the second method,
\citet{Yushkevich2009} hippocampal subfields were acquired and labelled on
high-resolution MRI data from post-mortem medial temporal lobe samples. Using
nonlinear registration guided by manually derived hippocampus masks and specific
landmarks, the authors demonstrated accurate parcellation of hippocampal
subfields in unlabelled MRI volumes.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and SNT \gs segmentations from the ADNI1 dataset \citep{Hsu2002}. This
form of cross-validation allows us to rigorously validate a large number of
parameter settings of \mb (atlas and template library sizes, registration
algorithm, and label fusion method) and select the best parameters to use in
subsequent experiments.

\paragraph{\adnidataset dataset} 
Data used in the preparation of this article were obtained from the Alzheimer’s
Disease Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). The ADNI
was launched in 2003 by the National Institute on Aging (NIA), the National
Institute of Biomedical Imaging and Bioengineering (NIBIB), the Food and Drug
Administration (FDA), private pharmaceutical companies and non-profit
organizations, as a \$60 million, 5-year public- private partnership. The primary
goal of ADNI has been to test whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological markers, and clinical and
neuropsychological assessment can be combined to measure the progression of mild
cognitive impairment (MCI) and early Alzheimer’s disease (AD).  Determination of
sensitive and specific markers of very early AD progression is intended to aid
researchers and clinicians to develop new treatments and monitor their
effectiveness, as well as lessen the time and cost of clinical trials.

The Principal Investigator of this initiative is Michael W. Weiner, MD, VA
Medical Center and University of California – San Francisco. ADNI is the result
of efforts of many co- investigators from a broad range of academic institutions
and private corporations, and subjects have been recruited from over 50 sites
across the U.S. and Canada. The initial goal of ADNI was to recruit 800 subjects
but ADNI has been followed by ADNI-GO and ADNI-2.  To date these three protocols
have recruited over 1500 adults, ages 55 to 90, to participate in the research,
consisting of cognitively normal (CN) older individuals, people with early or
late MCI, and people with early AD. The follow up duration of each group is
specified in the protocols for ADNI-1, ADNI-2 and ADNI-GO. Subjects originally
recruited for ADNI-1 and ADNI-GO had the option to be followed in ADNI-2. For
up-to-date information, see www.adni-info.org.

@@

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel dimensions of $1.25mm \times 1.25mm \times 1.2mm$.

@@

In this experiment, our dataset consists of 69 1.5T images and corresponding \gs
segmentations. In each validation round, the dataset is partitioned into a
training set consisting of images and \gs segmentations used as an atlas
library, and a validation set consisting of the remaining images segmented by
both \mb and multi-atlas. The computed segmentations were compared to the \gs
segmentations (see below).

A total of ten validation rounds were performed on each subject in the dataset,
over each combination of parameter settings. The parameter settings
explored are: atlas library size (1-9), template library size (1-20),
registration method (\ants or \animal, described below), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote, described below).  A total of $10 \times 69 \times 9
\times 20 \times 2 \times 3 = \num{7452000}$ validation rounds were conducted,
resulting in a total of $\num{1490400}$ segmentations analysed.

@@

The \animal algorithm carries out image registration in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using a blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller full width at half maximum (FWHM). The final transformation is a set of
local translations defined on a bed of equally spaced nodes that were estimated
through the optimization of the correlation coefficient.  For the purposes of
this work we used the regularization parameters optimized in
\citet{Robbins2004}, displayed in Table \ref{tab:ANIMAL-params}.

@@

\paragraph{Label fusion methods}
Label fusion is a term given to the process of combining the information from
several candidate labels for an image into a single labelling.  In this
experiment we explore three fusion methods: 
\begin{description}
  \item[Voxel-wise Majority Vote]
  Labels are propagated from all template library images to a target.  Each
  output voxel is given the most frequent label at that voxel location amongst
  all candidate labels.

@@

  \item[Normalised Mutual Information Weighted Majority Vote]
  This method is similar to cross-correlation weighted voting except that image
  similarity is calculated by the normalised mutual information score over the
  region of interest\citep{Studholme2001}.  The {\tt itk\_similarity} utility
  from the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is
  used to calculate the normalised mutual information measure between two images.
\end{description}

@@

\paragraph{Winterburn Atlases dataset} 
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
segmentations of five in-vivo $300\mu$ isotropic T1-weighted MR images. The
segmentations include subfield segmentations for the cornu ammonis (CA) 1; CA2
and CA3; CA4 and dentate gyrus; subiculum; and strata radiatum (SR, strata
lacunosum (SL), and strata moleculare (SM). Subjects in the Winterburn atlases
range in age from 29-57 years (mean age of 37), and include two males and three
females.

@@

In part 2a of this experiment, we have modified the LOOCV experimental
design in order to compare \mb segmentation accuracy with the accuracy of image
resampling error. The five $0.3mm$-isotropic voxel Winterburn atlas images and
labels are used as the atlas library, and LOOCV is conducted using the five Winterburn
atlas images subsampled (using trilinear interpolation) to
0.9mm-isotropic voxel resolution (referred to as the {\em Subsampled} dataset)
as input subjects. Subsampling the target images allows us to assess \mb in a
typical segmentation scenario (high-resolution atlases and lower-resolution
subjects).  The template library is composed of all five subject images, plus an
additional set of 3T T1 images (0.9mm-isotropic voxels) of healthy subjects
acquired separately (Table \ref{tab:WAval-healthy-demographics}). The optimal
size of template library, registration method, and label fusion method found in
Experiment 1 are used.  As with traditional LOOCV, each subject image is
segmented by \mb with that subject's image excluded from the atlas library.  In
part 2b of this experiment, we carry out a similar validation in which the
subject images are separately acquired T1 BRAVO images of four of the five
Winterburn atlas subjects (referred to as the {\em T1 BRAVO} dataset).  Figure
\ref{fig:LOOCV-schematic} illustrates the experimental set up of both parts of
this experiment.

@@

\paragraph{Evaluation method}  
To assess the \mb LOOCV segmentations we compute the relative percent error in 
hippocampal volume with respect to the full resolution Winterburn atlas 
segmentations. In addition, by computing the relative error in volume of the
Winterburn atlas labels resampled (with nearest-neighbour interpolation) to
0.9mm-isotropic voxels, we obtain a baseline error against which to assess \mb
segmentations.

@@

To validate that the \mb works effectively in the context of other neurological
disorders, in this experiment we use the Winterburn atlases to derive 
whole hippocampal segmentations of a dataset of patients with Schizophrenia. The
resulting segmentations are assessed for quality by comparison with expert
manual segmentations.

@@

Manual segmentation of each subject whole hippocampus is produced following a
validated segmentation protocol \citep{Pruessner2000}.

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2) ignoring subfield delineations.  All images from the
\FEPdataset dataset are segmented by \mb.  The optimal size of template library,
registration method, and label fusion method found in Experiment 1 are used.

\paragraph{Evaluation method}
The Pruessner and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated \citep{Winterburn2013}. This
difference poses a problem for evaluation by measuring overlap.  That is, since
different protocols will necessarily produce segmentations that do not perfectly
overlap, the degree of overlap cannot be solely used to compare segmentation
methods using different protocols. In place of an overlap metric, we can assess
the degree of correlation in average bilateral hippocampal volume of the
subjects produced by each method. Specifically, a Pearson correlation is used.

@@

To validate \mb segmentation quality with respect to other established automated
hippocampal segmentation methods, \mb was applied to large dataset from the ADNI
project and the resulting segmentations were compared to those determined by
\freesurfer, \fsl, \maper and the SNT \gs segmentation.

\paragraph{ADNI1 dataset revisited} 
The {\em \adnidataset} standardized dataset contains $1919$ images in total.
SNT and FreeSurfer hippocampal volumes for a subset of images were provided by
ADNI, along with quality control data for each \freesurfer segmentation
(guidelines described in \citep{Hartig2010}).  Clinical and demographic data for
the entire \adnidataset dataset are shown in Table
\ref{tab:ADNI1-Dataset-Demographics}.

@@

One author (MP) performed visual quality inspection for MAGeT and FSL
segmentations using similar quality control guidelines (if either hippocampus
was under or over segmented by $10mm$ or greater in three or more slices then
the segmentation did not pass).  Only images meeting the conditions of having
segmentations from all methods (SNT, MAPER, FreeSurfer, FSL, and \mb) and also
passing quality control inspection were included in the analysis.

\paragraph{Evaluation method}
As in Experiment 3, the SNT \gs and Winterburn hippocampal segmentation
protocols differ in the neuroanatomical features delineated, and so we 
assessed \mb by the degree of (Pearson) correlation of average hippocampal volume
across subjects. We also computed the correlation in hippocampal volume between
existing, established automated segmentation methods -- \fsl, \freesurfer, and
\maper, and SNT semi-automated segmentations. Additionally, we evaluate the
volume-related fixed and proportional biases in all segmentation methods using
Bland-Altman plots \citep{Bland1986}.

@@

In this experiment we conducted 10 rounds of \mb and multi-atlas segmentation of
of 69 subjects using atlas and template library sizes varying from 1-9 and 1-20
images respectively, two registration algorithms (\ants or \animal), and three
label fusion techniques (unweighted, cross-correlation, and normalised mutual
information weighted majority vote).  Hippocampal \mb-based segmentations using
both \animal and \ants registration algorithms demonstrate good overlap with SNT
\gs segmentations (maximum mean DSC of \Sexpr{round(mb.best.kappa, digits=2)}
when using 9 atlases, 19 templates, ANTS registration, and majority vote label
fusion); Figure \ref{fig:ADNI1-xval-k-mean}).  Qualitatively, both \animal and
\ants-based segmentations demonstrate trend overlap accuracy that increases with
the size of atlas library and template library. Improvement in accuracy plateaus
with template libraries larger than ten images.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<fig:ADNI1-xval-k-mean, cache=TRUE, dependson='ADNI1-xval-prep', fig.width=7, fig.height=7>>=
stderr <- function(x) sqrt(var(x)/length(x))
ggplot(subset(all_data_mean, (templates.mb*atlases) %% 2 == 1),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,
               fun.ymin=function (x) mean(x) - stderr(x),
               fun.ymax=function (x) mean(x) + stderr(x), 
               geom='errorbar',
               aes(y=k.mb, weight=1, group=as.factor(atlases)), width=0.5) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Number of Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  opts(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text = element_text(size = 8))
@
  \caption{Mean Dice's Similarity Coefficient of \mb segmentations relative to
  SNT \gs segmentations for 69 ADNI1 subjects vs. atlas and template library
  size, registration algorithm, and label fusion method. Error bars indictate
  standard error.
  \label{fig:ADNI1-xval-k-mean}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: "increase" in mean kappa over multi-atlas
\begin{figure}
<<fig:ADN11-xval-k-diff,cache=TRUE,dependson='ADNI1-xval-prep'>>=
ggplot(subset(all_data_mean, reg_method=="ANTS" & 
                method.mb=="Majority Vote" & (templates.mb*atlases) %% 2 == 1), 
       aes(x=templates.mb, y=k_diff, colour=as.factor(atlases))) + 
  #facet_grid(reg_method~method.y) + 
  stat_summary(fun.data=mean_cl_boot,geom='errorbar',
               aes(y=k_diff,colour=as.factor(atlases), width=0.2)) +
  stat_summary(fun.y=mean, geom="point", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  stat_summary(fun.y=mean, geom="line", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  #stat_smooth(method=lm, formula=y~log(x)) + 
  geom_hline(aes(alpha=0.5, yintercept=0), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,2)) + scale_y_continuous(breaks=seq(-1,1,by=0.01)) + 
  scale_colour_hue(name="Number of Atlases") +
  scale_alpha_continuous(guide = "none") + 
  scale_linetype_discrete(guide = "none") + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference in mean DSC" ) +
  opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Difference in mean Dice's Similarity Coefficient between \mb and
  multi-atlas segmentations relative to \gs (SNT) segmentations vs. atlas and
  template library size when using the \ants registration method, and
  majority-vote label fusion. Error bars indicate standard error.
  \label{fig:ADNI1-xval-k-diff}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

ggplot(subset(stats.tests, direction > 0 & stat == "Variance"), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  #facet_grid( . ~ stat) + 
  geom_line(size=1) +
  geom_point(size=3) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference of variance during validation (p-value)") +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  Variance of segmentation accuracy between \mb and multi-atlas segmentation is
  computed for each subject across all ten rounds of validation. Shown on the
  y-axis (scaled logarithmically) is the  p-value resulting from a t-test
  comparing the distribution of variances at each parameter setting
  (atlas/template library size). Only points where \mb mean variability is lower
  than multi-atlas are shown. Dashed lines indicate a p-value of 0.05 and 0.01.}
  \label{ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
